Predicting I/O-performance in HPC using Artificial Neural Networks

### 2016-07-07 Paper/paper.tex
Affiliation.
Julians überarbeitung.

%\classify{MSC?}
\author{I.M.~Scientist\footnote{\label{susu}South Ural State University} \and U.R.~Author\footnoteref{susu}}

@@

The prediction of file access times is an important part for the modeling of storage systems of super computers. These models can be used to develop analysis tools which support the integration of efficient I/O-behavior.\\
In this paper, we analyzed the access times of parallel storage system of a super computer.
Therefore, we measured file access times in various test series and developed different models for predicting access times. The evaluation showed that models utilizing artificial neural networks achieved about 30\% smaller average prediction errors than linear models. 
A phenomenon in the distribution of file access times is of particular interest:
File accesses with equal parameter values have several typical access times.\\
The steps in the magnitude between these typical access times can be explained with a different processing of the file accesses in the storage system.\\
We developed a method to quantify the significance of knowledge about the internal processing for the prediction of file access times and proved it to be essential.

@@

Tools are demanded that help users of HPC-facilities to implement efficient input/output (I/O) in their programs.
It is difficult to find the best access parameters and patterns due to complex parallel storage systems.
The processing of file accesses in a storage system can be viewed as a task that is sequentially propagated along a I/O-path in the storage system.
Starting at the invoking processor, the storage system is searching for the data, going further and further through the memory hierarchy, until all data is found, so it can be returned to the processor.

Currently, users have to optimize their programs for each system individually without much assistance.
To develop tools which support the implementation of efficient I/O a computational, models of the storage system are important.
For instance, a tool could estimate the I/O path and classify accesses as outliers, if they behave abnormally.
For single hard disk systems such a model can be derived analytically \cite{Ruemmler94anintroduction}; however, for the complex storage system of a super computer, these models become too difficult to configure \cite{DBLP:conf/npc/ZhangLZJC10}.

@@

Because of the strong linear correlation between access time and access size, the problem seems to fit linear models.
Our results, however, show that the relation of file access parameters to access time is not sufficiently represented by linear models.
ANNs achieve significantly better results than linear models.
Our analysis suggests that the I/O-path used by the storage system significantly influences the file access time.
Therefore it becomes key for a good model of access times to derive knowledge about I/O-paths.
Unfortunately I/O-paths are difficult to deal with, as it is unknown which path was used for a file access.

This paper is organized as follows:
Related work is provided in Section\,\ref{sec:related}.
In Section\,\ref{modeling_access_times}, we explain our analysis of file access times. 
First we develop a simplified model of the I/O-path. Then a method for approximating I/O-paths with derived classes from the error of a prediction model is proposed. At last we introduce a set of models for access time prediction.
Afterwards in Section\,\ref{evaluation}, we evaluate our analysis of the storage system by measuring access times in various test series, studying the obtained access times and the predictions of our models to them.
In the end Section\,\ref{conclusion} summarizes the paper suggests possible future work.

@@

Generally storage systems are modeled in two different ways for access time prediction, using white-box- or black-box-modeling \cite{Crume:2013:FML:2538542.2538561}.
\begin{itemize}
	\item \textbf{White-box-modeling}: The storage system itself is simulated. Details of hardware components like rotation speed of the magnetic disk in a hard drive are considered. The processing of a file access can be simulated in the model and the resulting access time is then used as prediction for the actual system.
	Processing and resulting performance can be analyzed in detail on the model.
	\item \textbf{Black-box-modeling}: The model abstracts from the real storage system. 
	System performance is imitated without consideration of its occurrence. %Zustandekommen
	This procedure corresponds to an emulation of the storage system.
	In contrast to the white-box-model, processing of file accesses can't be analyzed on the model itself.
\end{itemize}
The two ways of modeling are fundamentally different and have to be differentiated.

\subsection{White-box-modeling versus black-box-modeling}
For the in-depth analysis of reasoning for behavior of a storage system a white-box-model is desirable.
On the one hand the modeled system is represented in the model and can thus be examined, on the other hand these models can be very precise if modeled correctly \cite{Ruemmler94anintroduction}.
The problems of white-box-modeling are, however, as clear as its flaws; they have to be modeled individually for every system and modeling becomes quite intricate for single hard drives already \cite{Crume:2013:FML:2538542.2538561}.
To approach the complexity of white-box-modeling Ruemmler and Wilkes analyzed the relevance of different hard drive components for the model deviation, to save effort for insignificant parts \cite{Ruemmler94anintroduction}.
However, white-box-modeling is usually used for simple systems like a single hard drive. For these hard drives white-box-modeling is already very demanding, hence for the complex parallel storage system of a super computer it's not a feasible approach \cite{DBLP:conf/npc/ZhangLZJC10}.\\

Application of black-box-modeling is easier and more flexible as it's independent from the individual system.
Stochastic approaches coupled with data mining methods are mostly used for black-box-modeling; for example, a combination of regression trees support vector regression \cite{Dai:2012:SDP:2477169.2477214}, or selective bagging classification and regression trees \cite{DBLP:conf/npc/ZhangLZJC10}.

\subsection{Prediction of I/O-performance with ANNs}
Computability of ANNs was reasearched by Rojas \cite{Rojas:1996:NNS:235222} and Cybenko \cite{cybenko:mcss} they demonstrated possibility of modeling non linear systems. Cybenko also proved the \textit{universal approximation theorem} which states that feed-forward networks with sufficient complexity can approximate any continuous functions on compact subsets of $\mathbb{R}^n$.
Therefore ANNs should be sufficient for predicting I/O-performance; however, no statement can be made about the necessary structure of the network for this task.\\
Crume et al. developed a method using ANNs which exploits periodic patterns in sequences of file access times \cite{Crume:2013:FML:2538542.2538561}.
A Fourier analysis is used to determine the most important frequencies, which are then used as input information for the ANNs.
In a following publication they move away from Fourier analysis, but then use additional sine waves as input for the ANNs \cite{crumelatent}.
This seems to be a promising approach for access time prediction of single hard drives, where the rotational characteristics of the magnetic disk are an essential consideration. It is, however, difficult to extrapolate behavior from a single disk to a distributed storage system consisting of thousands of disks that utilizes several optimization mechanisms.

\subsection{I/O-performance prediction in HPC}
Performance analysis in HPC is a important task to examine the system for improvements in usage.
For instance Liu et al. simulated scheduling algorithms for research \cite{liu2011towards}. 
The simulation used the white-box-modeling tool DiskSim for prediction of occurring file accesses \cite{Bucy08thedisksim}.\\
There are a few simulators for parallel storage systems, for example, CODES\,\cite{cope2011codes} utilizes a scalable infrastructure to investigate relevant research issues such as the importance of burst-buffers.
PIOSimHD is a simulator able to replay (MPI) application traces on a generic I/O system \cite{kunkel2013simulating}.

Analytical and machine learning models for predicting performance are another choice to optimize performance.
Kunkel et al. utilized access time prediction with decision trees for varying parameterizations of ROMIO for access of non-contiguous data \cite{UMLTPTPONI15}. 
Instead searching for optimal parameters by testing, they were able to find good values through prediction of performance.

\section{Modeling I/O Access Times}
\label{modeling_access_times}
\subsection{Characteristics of the Data}
This approach allows for an in-depth analysis of system behavior for different use cases.
We used a synthetic benchmark in which identical measurements for random or sequential I/O are performed and the access time is measured individually.
The resulting timings are stored together with file access parameters in a file that is then used to build and validate the models.
The stored parameters are:
\begin{itemize}
%	\item \textbf{File-ID}: Identifies the invoked file. Brauchen wir ja nicht für die Analyse
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{Access type}: Differentiates reading and writing file accesses.
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time for performing the I/O.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed is not directly available.

\subsection{Model of the I/O-path}
The internal processing of a file access in the storage system can be viewed using the I/O-path which is the path from the invoking processor to the storage medium that contains the data (larger accesses may be distributed across multiple servers). 
Once the data has been found it is passed-through the levels of memory hierarchy to caches.
The resulting access time depends on the depth of this I/O-path because storage media further along the path are increasingly slower.
While the first levels in memory hierarchy (CPU caches) are extremely fast, the main memory is already slower; the same applies for the step into the parallel storage system, that is connected via network to the computer nodes.
%The trade-off between memory capacity and performance can be visualized with the memory hierarchy as in figure %\ref{src/2000px-ComputerMemoryHierarchy.png}.
%\fig{width=.7\textwidth}{src/2000px-ComputerMemoryHierarchy.png}{Memory hierarchy.}
Reading file accesses are more severely affected by varying depths of I/O-paths than writing file accesses, which are usually propagated to the disk drives.

\subsubsection{I/O-path for access time prediction}
\label{sec:path_for_pred}
Due to the exponential decay of processing speed in the hierarchy, file accesses with similar access parameters, but varying I/O-path, are differentiated by a step in the magnitude of access time.
As the access time is dominated by the slowest component along the I/O-path, a step in the measured access time occurs between to I/O-paths of diverging length.
In Figure\,\ref{src/plot_SizeSorted_log_read_seq.png}, measurements of reading file accesses can seen.
The different groups of measurements with equal access parameters (points with the same color) are clearly visible.
For a few access sizes, some clusters are observable, this step in the magnitude of access time and can be explained with varying I/O-paths (this figure will be discussed in further detail in the following chapter).
\fig{width=.47\textwidth}{src/plot_SizeSorted_log_read_seq.png}{Measurements of reading file accesses. Access sizes increase from left to right and all measurements with equal access parameter values have the same color.}

\subsection{Estimating the I/O path using error classes}
\label{sec:error_classes}
We utilize a concept of error classes.
Each class is supposed to approximate an I/O-path using the available measurements of file access times.
To derive such knowledge, we need a characteristic that is representative for the different I/O-paths. 
Knowing the file access parameters and the approximation of I/O-path should allow us to predict the magnitude of access time for this process.\\
At a first thought, one could come up with the idea of clustering the data seen in Figure \ref{src/plot_SizeSorted_log_read_seq.png}, because the groups of measurements with varying I/O-paths might be correctly differentiated by it. 
This, however, is not a good approach, one would have to cluster for every set of measurements with equal access parameters individually, otherwise the clusters would be meaningless. 
In a real application scenario, a large number of different access parameters with only very few instances occur, which makes this approach unfeasible.\\
Another idea might be then to use the difference of an arbitrary value to the measured file access time as parameter for the clustering. 
Such a value would not be appropriate for all magnitudes of access time occurring in the data, I/O-paths differentiating measurements with small access time might not be found with this approach, because the gap between them doesn't matter at the scale of the chosen arbitrary value.\\
Our approach uses the residues of a model like linear regression that can't differentiate between measurements with equal access parameter values.
Basically the linear regression predicts an average access time for these measurements.
The deviation of an individual measurement for these access parameters to the predicted average value is then characteristic for the I/O-path.
Clustering the residues with a k-Means algorithm allows us then to assign each measurement to a specific cluster, which is the approximation of its I/O-path.
We call the retrieved clusters \textbf{error classes}.
We use the absolute error of linear regression as clustering parameter so that one cluster can, for example, represent an deviation of 1 millisecond.
Ideally it might be the case that file accesses independent of the access parameters that took 1 millisecond longer than usual were processed on the same I/O-path.
In that case error classes are directly corresponding to I/O-paths. \medskip

We will use error classes to quantify the importance of knowledge about I/O-paths for access time prediction. 
For the actual prediction of access times error classes can't be used, because a measured access time is required to determine the error class of a measurement.
However, the method could be used for a tool, which analyses the I/O made during application of a program. 
It could analyze whether more file accesses throughout execution were slower than expected and therefore used unfavorable I/O-paths.
If a direct correlation of error class to I/O-paths were found, such an tool might be able to inform about percentages of used I/O-paths.

\subsection{Models}
Access times of file accesses can be predicted using various models.
While we evaluated several models for this paper, only a few relevant models are described; more models are described in \cite{VVEIHUDVVN15}. 
% TODO: Soll ich meine Arbeit noch woanders zitieren?
Each model seeks to correlate its known file access parameters to the corresponding access time. 
Linear regression is used as a baseline model with a simple mapping of access size to access time.
Additionally, three models with different input information utilizing ANNs were used:
\begin{itemize}
	\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
	\item A second ANN-model, called \textbf{ema-model}, received additionally to the parameters of the previous model information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance.
	\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
\end{itemize}\medskip

The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.\\
Figure\,\ref{periodic} shows a small series of measurements of sequentially read file accesses.
The peaks in access time are occurring periodically in this particular sequence. Such behavior appears only partially, often access times fluctuate without observable periodicity as in Figure\,\ref{sporadic}.
An occurrence of a periodic pattern could be explained with a processing of the storage system in which a larger amount of data is loaded into the cache at the same time as soon as a cache-miss happens.
\begin{figure}[!h]
	\centering
	\begin{minipage}[b]{0.47\textwidth}
		\includegraphics[width=\textwidth]{src/plot_From100001to100250_read_seq.png}
		\caption{Small sequence of reading file accesses with periodic peaks in access time.}
		\label{periodic}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.47\textwidth}
		\includegraphics[width=\textwidth]{src/plot_First250_read_seq.png}
		\caption{Small sequence of reading file accesses with sporadic peaks in access time.}
		\label{sporadic}
	\end{minipage}
\end{figure}
%\fig{width=0.47\textwidth}{src/plot_From100001to100250_read_seq.png}{Small sequence of reading file accesses with periodic peaks in access time.}
%\fig{width=0.47\textwidth}{src/plot_First250_read_seq.png}{Small sequence of reading file accesses with sporadic peaks in access time.}
With knowledge about these period of peaks a model could make better access time predictions.
Our second model is supposed to use its input information of previous data throughput for that.
The data throughput has peaks the same period as the access time.
Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
\begin{equation}
EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
\end{equation}
With $t(i)$ the access time of measurement $i$.
The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
In Figure \ref{src/ema2.png} the EMA-function to a graph with periodic peaks can be seen.
\fig{width=.47\textwidth}{src/ema2.png}{Exponential moving average in red for a graph with perdiodic peaks.}\\
The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions on sequences with periodic peaks in access time. Depending on how relevant periodic behavior of access times is in our measurements, this additional input information could improve the model.\medskip

@@

\section{Evaluation}
\label{evaluation}
The execution and results of analysis are summarized in the following.
First the test system used for measurements and the strategy for systematic measurement are presented.
Then follows the study of received data and computed error classes.
And in the end the quality of predictions of the different models are examined.

\subsection{Test system}
The measurements were done on the super computer Mistral of the DKRZ (Deutsches Klimarechenzentrum).
Mistral operates with the parallel distributed file system Lustre (Lustre 2.5, Seagate's edition) and consists of over 1500 computing nodes and 30 petabyte storage with a memory bandwidth of 300\,GiB/s.
Each computing node consists of two E5-2680v3 with a clock rate of 2.5\,GHz and 30\,MiB L3 Cache.
Measurements were done during daily operation, typical fluctuations in workload may have influenced the execution of benchmarks.

\subsection{Benchmarking}
We created a simple benchmark (io-model) that used POSIX read/write() and measured time for each I/O individually.
In this paper, it is executed on one node with one processor to study the quality of the predictions.
Each series of measurements follows a certain pattern to study system behavior systematically.\\
Sequential and random I/O patterns are measured, each with reading and writing file accesses.
Our test file has a size of 10\,GiB which doesn't fit in the main memory of computer nodes, also measurements from  suggest that caching is not effective for these measurements (sequential reads are improved by read-ahead, though).
Occasional access of hard drives in the parallel storage system is likely.
Most series of measurements contain 10\,000 file accesses with a certain access size, only series with sequential access and an access size of greater then 2\,MiB have fewer measurements as the end of our 10\,GiB file is reached.
Every series with specific access pattern, access size and access type is repeated three times.
Access sizes are varying from 1\,B to 16\,MiB (in detail:  1\,B, 4\,B, 16\,B, 64\,B, 256\,B, 1\,KiB, 4\,KiB, 8\,KiB, 16\,KiB, 64\,KiB, 256\,KiB, 512\,KiB, 1\,MiB, 2\,MiB, 4\,MiB, 8\,MiB and 16\,MiB).

\subsection{Analysis of measurements}
\label{sec:measurements}
At first the measured access times for the four different cases are summarized in Table\,\ref{overview}.
\tab{overview}{Overview of file access times for the different cases.}{
	\centering
	\scriptsize
	\begin{tabular}{|r|r|r|r|r|r|r|}\hline%
		Case & Min. value  & 1. Quartile & Median & Arith. mean & 3. Quartile & Max. value \\\hline\hline
		\csvreader[late after line=\\\hline]%
		{src/data_summary.csv}{Attribut=\Attribut,Min=\Min,Quartil1=\L, Median = \Median, Mittel = \Mittel,Quartil3 = \Q, Max = \Max, Korrelation = \Korrelation}%
		{\Attribut & \Min & \L & \Median & \Mittel & \Q & \Max}%
	\end{tabular}
}\\
As expected, sequential file accesses are on average much faster than random accesses especially for the reading case, also random read are slower than writes.\\
In Figure \ref{correlations}, we consider the correlations between file access parameters and the resulting access time, a strong correlation can be exploited for access time prediction.
\tab{correlations}{Correlations between file access parameters and access time.}{
	\centering
	\scriptsize
	\begin{tabular}{|r|r|r|}\hline%
		Attribute & Sequential access pattern & Random access pattern\\\hline\hline
		\csvreader[late after line=\\\hline]%
		{src/data_summary_korrelationen.csv}{Attribut=\Attribut,seq = \seq, rnd = \rnd}%
		{\Attribut & \seq &\rnd}%
	\end{tabular}
}\\
It is to note that sequential access leads to a constant delta-offset of $0$.
Sequential file accesses have with 97.3\% a clear correlation to access time, this is with only 32.91\% to a lesser degree still true for the random accesses.
This is why linear regression can be expected to have acceptable results for access time prediction, especially for the sequential access case.\medskip

>>>>

%\classify{MSC?}
\author{Fabian Schmidt\footnote{\label{susu}Universität Hamburg, E-Mail:\url{2schmid@informatik.uni-hamburg.de}} \and Julian M. Kunkel\footnote{German Climate Computing Center (DKRZ)}}

@@

The prediction of file access times is an important part for the modeling of supercomputer's storage systems. 
These models can be used to develop analysis tools which support the users to integrate efficient I/O-behavior.\\
In this paper, we analyze and predict the access times of a Lustre file system from the client perspective.
Therefore, we measure file access times in various test series and developed different models for predicting access times. 
The evaluation shows that models utilizing artificial neural networks achieved about 30\% smaller average prediction errors than linear models. 
A phenomenon in the distribution of file access times is of particular interest:
File accesses with identical parameters show several typical access times.\\
The typical access times usually differ by orders of magnitude times and can be explained with a different processing of the file accesses in the storage system -- an alternative I/O path.\\
We investigate a method to automatically determine the alternative I/O path and quantify the significance of knowledge about the internal processing.
It is shown that the prediction error is improved significantly with this approach.

@@

Tools are demanded that help users of HPC-facilities to implement efficient Input/Output (I/O) in their programs.
It is difficult to find the best access parameters and patterns due to the complexity of parallel storage systems.
The processing of file accesses in a storage system can be viewed as a task that is sequentially propagated along a I/O-path in the storage system.
Starting at the invoking processor, the storage system is searching for the data, going further and further through the memory hierarchy, until all data is found, so it can be returned to the processor.

Currently, users have to optimize their programs for each system individually without much assistance.
To develop tools which support the implementation of efficient I/O a computational, models of the storage system are important.
For instance, a tool could estimate the I/O path and classify accesses as outliers, if they behave abnormally.
For single hard disk systems such a model can be derived analytically~\cite{Ruemmler94anintroduction}; however, for the complex storage system of a supercomputer, these models become too difficult to configure~\cite{DBLP:conf/npc/ZhangLZJC10}.

@@

Because of the strong linear correlation between access time and access size, the problem seems to fit linear models.
However, our results show that the relation of file access parameters to access time is not sufficiently represented by linear models.
ANNs achieve significantly better results than linear models.
Our analysis suggests that the I/O-path used by the storage system significantly influences the file access time.
Therefore it becomes key for a good model of access times to derive knowledge about I/O-paths.
Unfortunately I/O-paths are difficult to deal with, as it is unknown which path was used for a file access.

This paper is organized as follows:
Related work is provided in Section\,\ref{sec:related}.
In Section\,\ref{modeling_access_times}, we explain our analysis of file access times. 
Firstly, we develop a simplified model of the I/O-path. 
Next, a method for approximating I/O-paths with derived classes from the error of a prediction model is proposed. 
At last, we introduce a set of models for access time prediction.
Afterwards in Section\,\ref{evaluation}, we evaluate our analysis of the storage system by measuring access times in various test series, studying the obtained access times and the predictions of our models to them.
Finally, Section\,\ref{conclusion} summarizes the paper and suggests future work.

@@

Generally, storage systems are modeled in two different ways for access time prediction; either using white-box modeling or black-box modeling~\cite{Crume:2013:FML:2538542.2538561}.
\begin{itemize}
	\item \textbf{White-box modeling}: The storage system itself is simulated. 
	Details of hardware components like rotation speed of the magnetic disk in a hard drive are considered. 
	The processing of a file access can then be simulated in the model and the resulting access time is then used as prediction for the actual system.
	Processing and resulting performance can be analyzed in detail on the model.
	\item \textbf{Black-box modeling}: The model abstracts from the real storage system. 
	System performance is approximated without considering the causes. %Zustandekommen
	This procedure corresponds to an emulation of the storage system.
	In contrast to the white-box model, processing of file accesses can't be analyzed on the model itself.
\end{itemize}
These two ways of modeling are fundamentally different and have to be differentiated.

\subsection{White-box-modeling versus black-box-modeling}
For the in-depth analysis of reasoning for behavior of a storage system, a white-box-model is desirable.
On the one hand, the modeled system is represented in the model and, thus, can be examined. 
On the other hand, these models can be very precise if modeled correctly \cite{Ruemmler94anintroduction}.
There are important limitations of white-box modeling, however. 
For every system an individual model has to be created and a model becomes quite intricate for single hard drives already \cite{Crume:2013:FML:2538542.2538561}.
To approach the complexity of white-box modeling, Ruemmler and Wilkes analyzed the relevance of different hard drive components for the model deviation, to save effort for insignificant parts \cite{Ruemmler94anintroduction}.
However, white-box modeling is usually used for simple systems like a single hard drive. For these hard drives white-box-modeling is already very demanding, hence for the complex parallel storage system of a supercomputer it's not a feasible approach \cite{DBLP:conf/npc/ZhangLZJC10}.\\

Application of black-box-modeling is easier and more flexible as it's independent from the individual system.
Stochastic approaches coupled with data mining methods are mostly used for black-box modeling; for example, a combination of regression trees support vector regression \cite{Dai:2012:SDP:2477169.2477214}, or selective bagging classification and regression trees \cite{DBLP:conf/npc/ZhangLZJC10}.

\subsection{Prediction of I/O-performance with ANNs}
Computability of ANNs was researched by Rojas \cite{Rojas:1996:NNS:235222} and Cybenko \cite{cybenko:mcss}; they demonstrated possibility of modeling non-linear systems. 
Cybenko also proved the \textit{universal approximation theorem}, which states that feed-forward networks with sufficient complexity can approximate any continuous functions on compact subsets of $\mathbb{R}^n$.
Therefore ANNs should be sufficient for predicting I/O-performance as long as there are no contradictions in behavior. 
However, no statement can be made about the necessary structure of the network for this task.\\
Crume et al. developed a method using ANNs which exploits periodic patterns in sequences of file access times \cite{Crume:2013:FML:2538542.2538561}.
A Fourier analysis is used to determine the most important frequencies, which are then used as input information for the ANNs.
In a following publication \cite{crumelatent} they move away from Fourier analysis, but then use additional sine waves as input for the ANNs.
This seems to be a promising approach for predicting access times of single hard drives, where the rotational characteristics of the magnetic disk are an essential consideration. 
It is, however, difficult to extrapolate behavior from a single disk to a distributed storage system consisting of thousands of disks that utilizes several optimization mechanisms.

\subsection{I/O-performance prediction in HPC}
Performance analysis in HPC is a important task to examine the system for improvements in usage.
For instance Liu et al. simulated scheduling algorithms for research \cite{liu2011towards}. 
The simulation used the white-box modeling tool DiskSim for prediction of occurring file accesses \cite{Bucy08thedisksim}.\\
There are a few simulators for parallel storage systems, for example, CODES\,\cite{cope2011codes} utilizes a scalable infrastructure to investigate relevant research issues such as the importance of burst-buffers.
PIOSimHD is a simulator able to replay (MPI) application traces on a generic I/O system \cite{kunkel2013simulating}.

Analytical and machine learning models for predicting performance are another choice to optimize performance.
Kunkel et al. utilized access time prediction with decision trees for varying parameterizations of ROMIO for access of non-contiguous data \cite{UMLTPTPONI15}. 
Instead searching for optimal parameters by testing, they were able to find good values through estimating performance.

\section{Modeling I/O Access Times}
\label{modeling_access_times}
\subsection{Characteristics of the Data}
We used a synthetic benchmark in which identical measurements for random or sequential I/O are performed and the access time is measured for each operation individually.
The resulting timings are stored together with file access parameters in a file that is then used to build and validate the models.
This approach allows for an in-depth analysis of system behavior for different use cases from the perspective of a single client.
The stored parameters are:
\begin{itemize}
%	\item \textbf{File-ID}: Identifies the invoked file. Brauchen wir ja nicht für die Analyse
	\item \textbf{Access size}: Number of bytes to read or write.
	\item \textbf{Access type}: Differentiates reading and writing.
\end{itemize}
Directly measurable or derivable attributes are:
\begin{itemize}
	\item \textbf{Offset}: Distance of file beginning to the starting point of access.
	\item \textbf{Delta-Offset}: Can be calculated for file access $i$ as Offset[$i$] - (Offset[i-1] + access size[i-1]).
	\item \textbf{Access time}: Time for performing the I/O.
\end{itemize}
Knowledge of internal aspects of the system about the current system utilization or about the storage media that have to be addressed is not utilized in the model.

\subsection{Model of the I/O-path}
The internal processing of a file access in the storage system can be viewed using the I/O-path, which is the path from the invoking processor to the storage medium that contains/will store the data (larger accesses may be distributed across multiple servers). 
The data read/written it is passed through the levels of the memory hierarchy.
The resulting access time depends on the depth of this I/O-path because storage media further along the path are increasingly slower.
While the first levels in memory hierarchy (CPU caches) are extremely fast, the main memory is already slower; the same applies for the step into the parallel storage system, that is deployed as servers connected via network to the computer nodes.
%The trade-off between memory capacity and performance can be visualized with the memory hierarchy as in figure %\ref{src/2000px-ComputerMemoryHierarchy.png}.
%\fig{width=.7\textwidth}{src/2000px-ComputerMemoryHierarchy.png}{Memory hierarchy.}
Reading file accesses are more severely affected by varying depths of I/O-paths than writing file accesses, which can be deferred using write-behind and propagated lazily to the disk drives.

\subsubsection{I/O-path for access time prediction}
\label{sec:path_for_pred}
Due to the exponential decay of processing speed in the hierarchy, file accesses with similar access parameters, but varying I/O-path, can be easily differentiated by a step in the magnitude of access time.
As the access time is dominated by the slowest component along the I/O-path, a step in the measured access time occurs between the  I/O-paths of diverging length.
In Figure\,\ref{src/plot_SizeSorted_log_read_seq.png}, measurements of sequential read accesses are shown.
The different groups of measurements with equal access parameters (points with the same color) are clearly visible.
For a few access sizes, some clusters are observable, this step in the magnitude of access time and can be explained with varying I/O-paths (this figure will be discussed in further detail in the following chapters).
\fig{width=.47\textwidth}{src/plot_SizeSorted_log_read_seq.png}{Time series for sequential reads. Access sizes increase from left to right and all measurements with equal access parameter values have the same color.}

\subsection{Estimating the I/O path using error classes}
\label{sec:error_classes}
We utilize a concept that we call \textit{error classes}.
Each class is supposed to approximate an I/O-path based on the error (residuum) of the observed access times to a baseline model.
%To derive such knowledge, we need a characteristic that is representative for the different I/O-paths. 
The baseline model is constructed as a function of the file access parameters.\\

At a first thought, one could come up with the idea of simply clustering the data seen in Figure~\ref{src/plot_SizeSorted_log_read_seq.png}, because the groups of measurements with varying I/O-paths might be correctly differentiated by it. 
This, however, is not a good approach, one would have to cluster for every set of measurements with identical access parameters individually as access time is strongly dependent on the access size. 
In a real application scenario, a large number of different access parameters with only very few instances occur, which makes this approach unfeasible.\\
Another idea might be then to use the difference of an arbitrary value to the measured file access time as parameter for the clustering. 
Such a value would not be appropriate for all magnitudes of access time occurring in the data.
I/O-paths that differ by little might not be found with this approach, because the gap between them doesn't matter at the scale of the chosen value.

Our approach uses the residues of a model like linear regression that can't differentiate between measurements with equal access parameter values.
Basically the linear regression predicts an average access time for the measurements of any given size.
The deviation of an individual measurement for these access parameters to the predicted average value is then characteristic for the I/O-path.
Clustering the residues with a k-Means algorithm allows us then to assign each measurement to a specific cluster, which is the approximation of its I/O-path.
We call the retrieved clusters \textbf{error classes}.
We use the absolute error of linear regression as clustering parameter so that one cluster can, for example, represent an positive deviation of 1 millisecond.
Ideally, it might be the case that file accesses independent of the access parameters that took 1 millisecond longer than usual are processed on the same I/O-path.
In that case error classes are directly corresponding to I/O-paths. 
However, this has to be investigated.

We use error classes to quantify the importance of knowledge about I/O-paths for access time prediction. 
For the actual prediction of access times, error classes can't be used, because a measured access time is required to determine the error class of a measurement.
However, the method could be used for a tool, which analyses the execution time of I/O on the client side made during application of a program and assign the observed time to the error class and, thus, the I/O path. 
It could analyze whether more file accesses throughout execution were slower than expected and, therefore, used unfavorable I/O-paths.
If a direct correlation of error class to I/O-paths were found, such an tool might be able to inform about percentages of used I/O-paths.

\subsection{Models}
Access times of file accesses can be predicted using various models.
While we evaluated several models for this paper, only a few relevant models are described; more models are described in \cite{VVEIHUDVVN15}. 
% TODO: Soll ich meine Arbeit noch woanders zitieren?
Each model seeks to correlate its known file access parameters to the corresponding access time. 
Linear regression is used as a baseline model with a simple mapping of access size to access time.
Additionally, three models with different input information utilizing ANNs are used:
\begin{itemize}
	\item The most simple ANN-model only received information about access sizes, delta-offsets and access types as input.
	\item A second ANN-model, called \textbf{ema-model}, received additionally to the parameters of the previous model information about the past data throughputs of the system, which can be used to exploit time dependencies of the I/O-performance.
	\item The third ANN-model, called \textbf{error-class-model}, received error classes in addition to the access parameters of the first ANN-model.
\end{itemize}\medskip

@@

The idea of the ema-model is to exploit periodic changes in the performance of the storage system.
Our analysis of measurements as well as the work of Crume et al. \cite{Crume:2013:FML:2538542.2538561} indicate that periodic phenomenons of access times in sequences of files accesses can be exploited for access time prediction.\\
Figure\,\ref{periodic} shows a small series of measurements of sequentially read file accesses.
The peaks in access time are occurring periodically in this particular sequence. Such behavior appears only partially, often access times fluctuate without observable periodicity as in Figure\,\ref{sporadic}.
An occurrence of a periodic pattern could be explained with a processing of the storage system in which a larger amount of data is loaded into the cache at the same time as soon as a cache-miss happens.
\begin{figure}[!h]
	\centering
	\begin{minipage}[b]{0.47\textwidth}
		\includegraphics[width=\textwidth]{src/plot_From100001to100250_read_seq.png}
		\caption{Small sequence of read accesses with periodic peaks in access time.}
		\label{periodic}
	\end{minipage}
	\hfill
	\begin{minipage}[b]{0.47\textwidth}
		\includegraphics[width=\textwidth]{src/plot_First250_read_seq.png}
		\caption{Small sequence of read accesses with sporadic peaks in access time.}
		\label{sporadic}
	\end{minipage}
\end{figure}
%\fig{width=0.47\textwidth}{src/plot_From100001to100250_read_seq.png}{Small sequence of reading file accesses with periodic peaks in access time.}
%\fig{width=0.47\textwidth}{src/plot_First250_read_seq.png}{Small sequence of reading file accesses with sporadic peaks in access time.}
With knowledge about these period of peaks a model could make better access time predictions.
The ema-model is supposed to use its input information of previous data throughput for that.
The data throughput has peaks the same period as the access time.
Internally the model calculates the exponential moving average (EMA) of data throughput of measurement $i$ as:
\begin{equation}
EMA(i) = 0.5 \cdot t(i)+ 0.5 \cdot EMA(i-1)
\end{equation}
With $t(i)$ the access time of measurement $i$.
The influence of data throughput of a measurement is exponentially decreasing in the series of EMAs.
In Figure \ref{src/ema2.png} the EMA-function for a function with periodic peaks can be seen.
\fig{width=.47\textwidth}{src/ema2.png}{Exponential moving average (in red) for a function with periodic peaks.}\\
The ema-model can use $EMA(n-1)$ for the prediction of access time of measurement $n$.
If the model is able to find threshold values of the EMA-function after which another peaks follows, it can use this input information for better access time predictions on sequences with periodic peaks in access time. Depending on how relevant periodic behavior of access times is in our measurements, this additional input information could improve the model.\medskip

\section{Evaluation}
\label{evaluation}
The execution and results of analysis are summarized in the following.
First, the test system used for measurements and the strategy for systematic measurement are presented.
Next, we explain the computation of error classes.
Finally, the quality of predictions of the different models are examined.

\subsection{Test system}
The measurements were done on the supercomputer Mistral (phase\,1 system) of the DKRZ (Deutsches Klimarechenzentrum)\footnote{\url{http://www.vi4io.org/hpsl/2016/de/dkrz/start}}.
The phase\,1 system of Mistral operates with the parallel distributed file system Lustre (Lustre 2.5, Seagate's edition) and consists of over 1500 computing nodes and 30 Petabyte storage.
Each computing node consists of two E5-2680v3 with a clock rate of 2.5\,GHz and 30\,MiB L3 Cache.
Measurements were done during daily operation, typical fluctuations in workload may have influenced the execution of benchmarks.
